---
title: "Introduction to R"
format:
  html: 
    toc: true
  revealjs:
    output-file: "week1-r-intro-slides.html"
    theme: simple
    slide-number: true
    chalkboard: true
    preview-links: auto
    footer: "CSS 211 | UC San Diego"
---

## Goals of the lecture {.smaller}

::: {.incremental}
- Brief tooling.
- Why R? 
- Introduction to "base R". 
- Brief preview of the `tidyverse`. 
:::

::: {.fragment}
::: {.callout-note icon="false"}
Note that today's lecture will extend to Friday's *in-class lab slot*. Depending on whether we have time on Friday, we can also work on the *take-hom lab* in Friday's class.
:::
:::


## Tooling (briefly) {.smaller}

> One of the most frustrating parts of programming is **tooling**: getting your computer set up to actually *do* the stuff you want to learn about.

In this class, we'll be working with the [R programming language](https://www.r-project.org/) using a desktop IDE called [RStudio](https://posit.co/download/rstudio-desktop/). 

::: {.incremental}
- Links to download and install RStudio can be found [here](https://posit.co/download/rstudio-desktop/). 
- Follow the instructions: will include downloading and installing **R**.  
- To avoid other tooling headaches, we'll just be using *Canvas* for course management.
- We won't be relying on GitHub, but it's also very useful and important!
:::


## Why R? {.smaller}

There are *many* different programming languages. Why use R? 

::: {.incremental}
- R was specifically designed for *data analysis*.  
- R supports a number of *open-source packages* to make data analysis easier.
  - `ggplot`, `dplyr`, `lme4`. 
  - We'll be learning all about these in the course!
- Other CSS-relevant languages include [Python](https://www.python.org/) and [SQL](https://www.w3schools.com/sql/). 
  - Learning R is not *incompatible* with learning these!
:::


## Introduction to "base" R {.smaller}

> "Base" R just refers to the set of functions and tools available "out of the box", without using additional packages like `tidyverse`.

Base R includes (but is not limited to):

::: {.incremental}
- Basic mechanics like **variable assignment**. 
- Simple functions like `plot`, as well as core *types* like **vectors**.
- Statistical methods like `lm` and `anova` (which we'll discuss later). 
:::



## Variable assignment {.smaller}

> **Variables** allow us to *store information* (values, vectors, etc.) so we can use it again later.

Here, we create a variable called `account`, so we can add to it.

```{r basic1}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

# Our first R variable
account <- 20 ### assign value to variable
account + 25 ### add to variable
```

::: {.fragment}
::: {.callout-note icon="false"}
You can also use `=` for assignment, but `<-` is the R convention. (For most purposes in this course, it shouldn't matter which you use, and I sometimes mix them up!)
:::
:::

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Try adding different numbers to `account`. What do you think will happen if you add a string like `'CSS'`?
:::
:::


## Basic variable types {.smaller}

> Each variable has a certain type or `class`. 

You can *do different things* with different types of variables. For instance, you can't calculate the `mean` of multiple characters, but you can for `numeric` types.

| Type | What it is | Example |
|------|------------|---------|
| **numeric** | Numbers (integers & decimals) | `age <- 25`, `gpa <- 3.7` | 
| **character** | Text strings | `name <- "Alice"` | 
| **logical** | TRUE/FALSE values | `passed <- TRUE` |
| **integer** | Whole numbers only | `count <- 5L` | 
| **factor** | Categorical data | `grade <- factor("A")` |

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Use the `typeof()` or `class()` functions to check the type of different variables or values. 
:::
:::


## Basic Operations with Numeric Variables {.smaller}

`numeric` variables allow for a number of *arithmetic operations* (like a calculator).

```{r operations1}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
# Creating numeric variables
my_var <- 5


my_var + 1 ### Addition
my_var * 2 ### Multiplication
my_var / 2 ### Division
my_var ** 2 ### Exponentiation
```


## Vectors: Building Blocks of R {.smaller}

> A **vector** is a collection of elements with the same `class`. 

Vectors can be created with the `c(...)` function.

```{r v1}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
# Creating vector
my_vector <- c(25, 30, 32)
print(my_vector)
print(my_vector[1])
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Create your own vector, this time with `character` types in it. Try **indexing** into different parts of the vector.
:::
:::

## Working with vectors {.smaller}

Like **scalars**, numeric vectors can be manipulated mathematically.

```{r v1-math}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

my_vector + 1 ### add 1 to each element
my_vector * 5 ### Multiply each element by 5
my_vector + c(1, 2, 3) ### Add vector to another vector
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
What do you think will happen if you try to add two vectors of different lengths?
:::
:::


## Functions {.smaller}

> A **function** implements some operation; you can think of it as a *verb* applied to some input. 

In CSS, you'll often be using functions to *summarize* your data (like a vector).


```{r v1-function}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
# Creating vector
heights <- c(60, 65, 62, 70, 72, 73)

mean(heights)
median(heights)
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Create a vector containing possible *incomes*; then create the `mean` and `median` of this vector.
:::
:::



## Creating vectors from distributions (pt. 1) {.smaller}

In addition to creating vectors *by hand*, we can use functions to create **random vectors** by *sampling* from some distribution, e.g., a **normal distribution** (`rnorm(x, mean, sd)`).

```{r rand-example}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
#| fig-width: 5
#| fig-height: 3
# Creating vector
v_norm = rnorm(100, mean = 50, sd = 2)
hist(v_norm)
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Try changing the different parameters of `rnorm` and then plotting the resulting vector again using `hist`. What do you notice about changing the `mean` or `sd`? 
:::
:::

## Creating vectors from distributions (pt. 2) {.smaller}

There are also many types of **distributions** beyond normal distributions.

::: {.incremental}
- Uniform distributions: use `runif`. 
- Binomial distributions: use `rbinom`. 
- Poisson distributions: use `rpois`. 
- **Sampling** from these distributions (and visualizing them) is a helpful way to learn about different statistical distributions. 
:::

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Create a uniform distribution with `runif` with 100 values ranging from `2` to `3`. If you're not sure how to do this, use `?runif` to learn more about the function.
:::
:::


## Interim summary {.smaller}

So far, we've covered a number of **core topics** in base R. 

::: {.incremental}
- Assigning and working with *variables*. 
- Different *types* of variables. 
- Applying *functions* to variables. 
- Creating *vectors* and visualizing them with `hist`. 
- Sampling from *statistical distribution*. 
:::

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Any questions before we move on to creating **dataframes** and other kinds of plots?
:::
:::

## Dataframes {.smaller}

> The `data.frame` class is a "tightly coupled collection of variables"; it's also a **fundamental data structure** in R.

::: {.incremental}
- Like a matrix, but with *labeled columns* of the same length. 
- Each column corresponds to a *vector* of values (numbers, characters, etc.). 
- Supports many useful operations. 
- Analogous to `pandas.DataFrame` in `Python`!
:::

::: {.fragment}
::: {.callout-note icon="false"}
Note that once we move to the `tidyverse`, we'll be working with `tibbles`, which are *basically* like a `data.frame`.
:::
:::


## Creating a `data.frame` {.smaller}

::: {.incremental}
- A `data.frame` can be created using the `data.frame` function.
- Pass in labeled vectors *of the same length*.
:::

```{r df-example}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
df_example = data.frame(hours_studied = c(0, 2, 2, 3, 5, 8), test_score = c(70, 85, 89, 89, 94, 95))
head(df_example, 2)
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Try creating your own `data.frame` with custom columns. For example, one column could be `movie_title` and another could be your `rating` of that movie. Make sure the columns are the same length!
:::
:::


## Exploring  a `data.frame` {.smaller}

We can use functions like `nrow`, `head`, and `colnames` to learn about our `data.frame`.

```{r df-example2}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
print(nrow(df_example)) ### How many rows?
print(colnames(df_example)) ### Column names
print(head(df_example, 2)) ### First two rows
print(str(df_example)) ### Structure of data
print(summary(df_example)) ### Summary of each column
```

## Accessing individual columns {.smaller}

You can access individual *columns* using the `dataframe$column_name` syntax.

```{r df-columns}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
df_example$hours_studied ### Get vector
summary(df_example$hours_studied) ### Get summary of vector
```

## Filtering a `data.frame` {.smaller}

In base R, you can **filter** a `data.frame` using the `df[CONDITION]` syntax, where `CONDITION` corresponds to a `logical` statement.

```{r df-filter}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
df_example[df_example$hours_studied > 2, ]
```

::: {.fragment}
::: {.callout-note icon="false"}
In the `tidyverse`, we can use the handy `filter` function.
:::
:::


## Simple *bivariate* plots {.smaller}

Once you have *multiple vectors*, you can plot the relationship between them, e.g., using a simple **scatterplot**.

```{r scatter}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
#| fig-width: 7
#| fig-height: 4
plot(df_example$hours_studied, 
     df_example$test_score,
     xlab = "Hours Studied", 
     ylab = "Test Score",
     pch = 16, # Filled circles
     col = "blue")
```

## Calculating *correlations* {.smaller}

You can also *quantify* the relationship between variables, e.g., using a **Pearson's r correlation coefficient**.

```{r cor}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
cor.test(df_example$hours_studied, df_example$test_score)
```

## Working with missing data {.smaller}

Real data often contains **missing values**. R represents these as `NA` (Not Available). We'll discuss these in more detail next week, but here's a preview:


```{r missing}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
# Vector with missing data
survey_responses <- c(85, 92, NA, 78, NA, 88)
mean(survey_responses)                    # Returns NA!
mean(survey_responses, na.rm = TRUE)      # Remove NAs first: 85.75
```

## Working with missing data (pt. 2) {.smaller}

You can *remove* missing data by **filtering** the `data.frame`, using the syntax below and the `is.na` condition.

```{r missing2}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
# Vector with missing data
survey_responses[is.na(survey_responses) == FALSE]
```


## Putting it together: simulating data {.smaller}

So far, we've discussed a number of useful concepts in R:

::: {.incremental}
- Working with **vectors**. 
- Simulating random distributions using `rnorm`. 
- Creating `data.frame` objects and plotting or analyzing them.
:::


::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Now, let's *simulate data*. 

- First, use `rnorm` to create a random normal distribution of `parent_heights` (use parameters that seem reasonable to you). 
- Then, create a second variable called `child_heights` that's *related* to `parent_heights`, ideally with some *random error* added. (**Hint**: Think about how a regression line works.). 
- Put these variables in a `data.frame`.
- Finally, **plot** the relationship between those variables and calculate the correlation.
:::
:::

## Simulating data {.smaller}


```{r simulating}
#| echo: true
#| code-line-numbers: true
parent_heights = rnorm(100, 65, 3)
child_heights = parent_heights + rnorm(100, 0, 2)
df_heights = data.frame(parent_heights, child_heights)

plot(df_heights$parent_heights, 
     df_heights$child_heights,
     xlab = "Parent Height", 
     ylab = "Child Height",
     pch = 16, # Filled circles
     col = "blue")

cor(df_heights$parent_heights, df_heights$child_heights)

```

## A conceptual preview of the `tidyverse` {.smaller}

Next week, we'll discuss the `tidyverse`: a set of packages and functions developed to make data analysis and visualization in R easier.

This includes (but is not limited to):

::: {.incremental}
- Functions for [**transforming data**](https://nyu-cdsc.github.io/learningr/assets/data-transformation.pdf), e.g., `filter` or `mutate`. 
- Functions for **merging data**, like `left_join` or `inner_join`. 
- Functions for **visualizing data**, like `ggplot`. 
:::


## Lecture wrap-up {.smaller}

This course is not *primarily* about programming in R, but programming in R is a **foundational skill** for other parts of this course.

This lecture (and accompanying lab) is intended to give you more comfort with the following concepts:

::: {.incremental}
- Working with **variables** and different types of data. 
- Creating and working with **vectors**.  
- Simple **plotting**. 
- Working with `data.frame` objects.
:::




