---
title: "Introduction to CSS 211"
format:
  html: 
    toc: true
  revealjs:
    output-file: "week0-intro-slides.html"
    theme: simple
    slide-number: true
    chalkboard: true
    preview-links: auto
    footer: "CSS 211 | UC San Diego"
---

## Goals of the lecture {.smaller}

::: {.incremental}
- Introductions  
- What is CSS?
- What is this course for?
- Logistics
:::

## Who am I?  {.smaller}

- [Assistant Teaching Professor, Cognitive Science](https://seantrott.github.io/). 
- [Research interests](https://seantrott.github.io/research/): large language models (LLMs), language comprehension, Theory of Mind. 

::: {.content-visible when-format="revealjs"}
## Who are you?  {.smaller}

**Turn to someone near you and share:**

::: {.incremental}
- Your name
- Your research interests 
- What drew you to CSS
:::
:::

## What is CSS?  {.smaller}

> In a nutshell, Computational Social Science focuses on computational approaches to social science.

At UCSD, Social Sciences encompasses many disciplines:

::: {.incremental}
- Economics. 
- Political Science. 
- Cognitive Science.
- Sociology.
:::

## What is social science?  {.smaller}

**Social science** refers to a domain of study: *social phenomena*.

::: {.incremental}
- Encompasses many **scales**: individual human behavior, political trends, etc. 
- Involves multiple **methods**: qualitative interviews, statistical analysis, simulations.
:::

## What are computational methods for?  {.smaller}


Another way of asking this is: **how do computational methods help us learn what we want to know**?

::: {.incremental}
- Many research questions benefit from *large-scale quantitative analysis*.  
  - Measuring trends over time.
  - Comparing effects of policy interventions. 
  - Modeling the relationships between variables. 
- Computational methods help us do this efficiently and reliably.
:::



## What is this course for? {.smaller}

The goal of this course is to introduce students to **foundational concepts and methods** in statistical modeling. We will emphasize both **hands-on application** and **conceptual understanding**. 

Key concepts include:

::: {.incremental}
- Data wrangling, summarization, and visualization.  
- Modeling data and designing analyses.
- Interrogating the assumptions of an analysis or dataset.
:::

::: {.fragment}
Hands-on work will be done using the R programming language.
:::


## Learning outcomes {.smaller}

My goal is that by the end of this course, students will be able to:

- **Define** and **explain** key concepts in statistical inference and regression analysis.
- **Identify** appropriate visualizations and statistical methods for different kinds of research questions and datasets.
- **Implement** data wrangling, visualization, and analysis workflows in R.
- **Interpret** and **evaluate** results (visualizations, fit models, etc.) in the context of a research question.
- **Design** and **implement** a complete statistical analysis project from research question to interpretation. 


## Why R? {.smaller}


> [R](https://www.r-project.org/) is a programming language used to perform statistical analysis, wrangle data, make data visualizaitons, and more. 

R is certainly not the *only* way to analyze data, but it's a useful tool in your toolbox for a couple reasons.

::: {.incremental}
- R has excellent packages (`lme4`) for running analyses with mixed effects models. 
- R has the [`tidyverse`](https://www.tidyverse.org/), a collection of packages for writing really *clean* code to wrangle, summarize, and visualize data.  
- The `tidyverse` includes [`ggplot`](https://ggplot2.tidyverse.org/), one of the best plotting packages I've ever used.  
- R is integrated with platforms like [RStudio](https://posit.co/download/rstudio-desktop/), which enable creation of sharable code, websites, and slides (like these).
:::


## R: basics  {.smaller}

R can be used to perform simple arithmetic calculations:

```{r basic1}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

# Our first R calculation
x <- 1 + 1
print(paste("The answer is:", x))
```

It can also be used to calculate summary statistics, like a `mean` or standard deviation (`std`).


```{r basic2}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

sample_vector = c(2, 4, 6, 8)
print(paste("Mean: ", mean(sample_vector)))
print(paste("SD: ", sd(sample_vector)))
```

## R: simple histogram  {.smaller}

R can also be used to *visualize data*.

```{r hist}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

vector = rnorm(100, mean = 50, sd = 2)
hist(vector)
```

## R: simple scatterplot  {.smaller}

*Scatterplots* can be used to visualize relationships between data.

```{r scatter}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

v1 = rnorm(100, mean = 50, sd = 2)
v2 = v1 * 2 + rnorm(100, mean = 0, sd = 2)
plot(v1, v2)
```


## R: simple scatterplot  {.smaller}

*Scatterplots* can be used to visualize relationships between data.

```{r scatter_no_plot}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

v1 = rnorm(100, mean = 50, sd = 2)
v2 = v1 * 2 + rnorm(100, mean = 0, sd = 2)
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
What do you think the `v2` line of code is doing?
:::
:::

## Logistics and course structure  {.smaller}

Designing this course required some forethought. CSS is a *broad* discipline: what methods and content are useful for everyone?

My approach:

> Course content and assessments should reflect underlying **learning outcomes**.

::: {.incremental}
- **Define** and **explain** key concepts in statistical inference and regression analysis.
- **Identify** appropriate visualizations and statistical methods for different kinds of research questions and datasets.
- **Implement** data wrangling, visualization, and analysis workflows in R.
- **Interpret** and **evaluate** results (visualizations, fit models, etc.) in the context of a research question.
- **Design** and **implement** a complete statistical analysis project from research question to interpretation. 
:::

## Assessment strategy {.smaller}

::: {.incremental}
| Assessment | Percentage | Learning Outcome |
|------------|------------|------------------|
| Labs | 20% |Implementation, interpretation, evaluation |
| Concept quizzes | 20% |Defining, explaining|
| Midterm | 25% |Defining, interpreting, evaluating |
| Final project | 35% |Designing, identifying, implementing, evaluating |
:::

## Course policies {.smaller}

::: {.incremental}
- **Attendance**: Recommended, not required. 
- **Dropped quiz**: Your lowest quiz will be dropped from your grade.
- **Late assignments**: Generally not accepted except in cases of emergencies. 
  - If you have a *midterm scheduling conflict*, let me know *as soon as possible* so we can try to accommodate it. 
- **Getting help**: Office hours from the teaching team generally preferred to email. 
:::

## Note on AI usage {.smaller}

My research focuses on large language models (LLMs), which have seen incredible progress in recent years. I probably don't need to tell you how powerful ChatGPT can be!

A few points need to be made here:

::: {.incremental}
- Tools like ChatGPT will likely keep getting better, and we shouldn't ignore that.
- Doing CSS "in the wild" may sometimes benefit from using ChatGPT!
- Separately, CSS "in the wild" is not done in an insolated test-taking environment.
- It's also still important (I think) to learn the **fundamentals** .
  - I can use ChatGPT more effectively when I know *what I want to do*.
- Thus, our assessments combine:
  - Take-home work (more "ecologically valid").
  - Work requiring "endogenous" problem-solving (in-person midterm).
:::

## Final project {.smaller}

The final project will be done independently, and will involve **replicating** from start to finish a published analysis in the CSS domain of your choice. 

::: {.incremental}
- You have a lot of freedom here, but I recommend:
  - Thinking about this **early on**. Happy to point people to useful resources!
  - Choosing something that's **challenging** (but also doable!). 
- **Extensions** to existing work (e.g., new analyses) will be encouraged. 
- Your **deliverables** will be:
  - A final report (turned in via Canvas). 
  - A final presentation (delivered week 10).
:::


## Tooling, prerequisites, and tech setup {.smaller}

This course will involve quite a bit of programming in R.

::: {.incremental}
- Week 1 will cover **tooling**. You'll need R installed and RStudio. 
- No explicit prerequisites, though some programming background (like CSS bootcamp!) will be helpful.
:::

## Topics and schedule {.smaller}

Topics will be roughly as follows:

::: {.incremental}
- **Week 1**: Introduction to R and the RStudio environment. 
  - (Plus some philophy of science.)
- **Weeks 2-3**: Data wrangling and visualization. 
- **Weeks 4-5**: Deep dive into linear regression. 
- **Weeks 6-8**: Advanced statistical techniques.  
- **Weeks 9-10**: Best practices, final project work and presentations.
:::

## Welcome to CSS 211!
