---
title: "Data Wrangling in R"
format:
  html: 
    toc: true
  revealjs:
    output-file: "week2-r-wrangling-slides.html"
    theme: simple
    slide-number: true
    chalkboard: true
    preview-links: auto
    footer: "CSS 211 | UC San Diego"
---

## Goals of the lecture {.smaller}

::: {.incremental}
- What is *data wrangling* and why should we learn it? 
- Principles of **tidy data**. 
- Wrangling with the `tidyverse`.
:::



## What is data wrangling? {.smaller}

> **Data wrangling** refers to the processes of transforming or manipulating raw data into a useful format for downstream *analysis* and *visualization*.

In CSS, you'll find yourself needing to: 

::: {.incremental}
- *Import* data from various sources, in various formats. 
- *Clean* messy variables or *recast* them to different types. 
- *Filter* and *transform* data. 
- *Address* missing values.
- *Reshape* data between different formats. 
- *Join* multiple datasets together. 
:::

[These are all supported by the `tidyverse`! But what *is* **tidy data**?]{.fragment}


## Tidy data {.smaller}

[Wickham (2014)](https://www.jstatsoft.org/article/view/v059i10/0) defines **tidy data** as follows:

> "Tidy datasets are easy to manipulate, model and visualize, and have a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table."

Let's address each of these in turn:

::: {.incremental}
1. Each *variable* is a column. 
2. Each *observation* is a row. 
3. Each *type of observational unit* is a table.
:::

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
With a partner, discuss what each of these criteria might mean.
:::
:::

## Each variable is a *column* {.smaller}

> A **variable** contains values that measure the same attribute across units.

::: {.incremental}
- *Examples*: `height`, `age`, `income`, `test_score`, `population`.
- Each variable gets its own *column*.
:::

::: {.fragment}
**Untidy example:**

| country | gdp_2020 | gdp_2021 | gdp_2022 |
|---------|----------|----------|----------|
| USA     | 21.35   | 23.368    | 26.01  |
| China   | 14.72    | 18.2    | 18.32   |
| Germany | 3.94     | 4.35     | 4.16    |
:::

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Why is this *not tidy*? How would you make it tidy?
:::
:::

## Each variable is a *column* {.smaller}

Now, we have a *column* for `country`, `year,` and `gdp`.

::: {.fragment}
**Tidy example:**

| country | year | gdp_trillion |
|---------|------|--------------|
| USA     | 2020 | 21.35        |
| USA     | 2021 | 23.368       |
| USA     | 2022 | 26.01        |
| China   | 2020 | 14.72        |
| China   | 2021 | 18.2         |
| China   | 2022 | 18.32        |
| Germany | 2020 | 3.94         |
| Germany | 2021 | 4.35         |
| Germany | 2022 | 4.16         |
:::

## Each observation is a *row* {.smaller}

> An **observation** contains all values measured on the same *unit* across attributes.

::: {.incremental}
- One row = one complete case or measurement. 
- No splitting observations across multiple rows.
:::

::: {.fragment}
**Untidy example:**

| state      | measure          | value |
|------------|------------------|-------|
| California | population       | 39.5  |
| California | electoral_votes  | 54    |
| Texas      | population       | 31  |
| Texas      | electoral_votes  | 40    |
:::

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Why is this *not tidy*? How would you make it tidy?
:::
:::

## Each observation is a *row* {.smaller}

`population` and `electoral_votes` are separate *measures* for the same *case* (observation), so should be in the same row.

::: {.fragment}
**Tidy version:**

| state      | population_millions | electoral_votes |
|------------|--------------------|-----------------| 
| California | 39.5               | 54              |
| Texas      | 31.0               | 40              |
:::


## Each type of observational unit is a *table* {.smaller}

> Different types of things being measured should be in separate tables.


::: {.incremental}
- *Example*: Record student test scores in one table, and school-level information in another. 
- *Example 2*: Record monthly temperatures for cities in one table, and city *altitude* in another. 
- This avoids repeating information (like school name for every student). 
- Once we're ready, we can **join** these tables. 
:::

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Can you think of other examples of "observational units" you would want to keep in separate tables?
:::
:::

## Tidy data revisited {.smaller}

> "Tidy datasets are easy to manipulate, model and visualize, and have a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table." (Wickham, 2014)

::: {.incremental}
- These guidelines aren't *arbitrary*.
- A **standard tidy structure** is helpful for *visualizing* and *analyzing* data. 
:::

::: {.fragment}
::: {.callout-note icon="false"}
In fact, I often conceptualize "tidy" data in terms of:

- What are the *axes* in a plot I want to make? (These should be separate *columns*.)
- What are the *terms* in a regression formula I want to write? (These should be separate *columns*.)
:::
:::

## Welcome to the `tidyverse` {.smaller}

> The [`tidyverse`](https://www.tidyverse.org/) is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures."

::: {.incremental}
- Fundamentally, the `tidyverse` rests on the philosophy of **tidy data**. 
- Includes a range of functions, from *importing data* (`read_csv`) to *merging datasets* (`inner_join`). 
:::

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Installation Help

- Install the `tidyverse` using `install.packages("tidyverse")`. 
- Then, *load* it using `library(tidyverse)`. 
:::
:::


## Loading the `tidyverse` {.smaller}

```{r loading}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment
### Loading the tidyverse
library(tidyverse)
tidyverse::tidyverse_packages()
```

::: {.incremental}
- We'll cover some of these packages later in this class, like `broom` and `ggplot2`. 
- This week, we'll focus on functions from `readr`, `magrittr`, `dplyr`, and  `tidyr`.
:::


## The pipe (`%>%`) {.smaller}

> The **pipe operator** (`%>%`) lets you chain functions together in a readable way.

::: {.incremental}
- Takes the output from the left side and feeds it as the first argument to the right side. 
- Instead of **nesting** functions, can **pipe** outputs into each other. 
- Makes code more readable and easier to follow. 
- We'll see examples coming up!
:::


## Part 1: Basic data manipulation

Importing, filtering, mutating, and summarizing data.


## Reading in a dataset {.smaller}

One of the most fundamental things you'll need to do is *import* data, e.g., from a `.csv` file. 

- `read_csv` takes as input a *filepath* (either locally, or a URL). 
- Output is a `tibble`. 

```{r read_csv}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_conc <- read_csv("https://raw.githubusercontent.com/seantrott/ucsd_css211_datasets/main/main/wrangling/concreteness.csv")
nrow(df_conc)
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in
Use of the functions we've discussed (`nrow`, `names`, `head`, etc.) to explore the dataset.
:::
:::

## Basics of `dplyr` {.smaller}

> `dplyr` is a package in the `tidyverse` that contains functions for *transforming* and *manipulating* dataframes.

Each function is a "verb", for instance:

::: {.incremental}
- `filter` relevant rows. 
- `mutate` the dataset by adding new variables. 
- `pull` specific columns by name.  
- `rename` specific columns. 
- `group_by` specific values or factors. 
- `summarize` dataset with summary statistics. 
:::

## The `filter` {.smaller}

> `filter` works by selecting a *subset* of rows based on some **boolean condition**. 

```{r filter}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_conc %>%  ### piping!
  filter(Word == "class")
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

Try using `filter` to return words that:

- Have a `Concreteness > 4.5`. 
- Have a `Dom_Pos` of `Noun`. 

**Hint**: You can *chain* multiple filter statements together with a `%>%` operator.
:::
:::

## `mutate` your data {.smaller}

> `mutate` works by *creating* a new variable (or *overwriting* an existing one) via applying some transformation to an existing variable or set of variables.

Let's create a **log frequency** variable, since `Frequency` is highly right-skewed.

```{r mutate1}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_conc = df_conc %>%  ### piping!
  mutate(log_freq = log10(Frequency))

summary(df_conc$log_freq)
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

Why do you think the *mininum* `log_freq` value is `-Inf`?
:::
:::

## `mutate` (pt. 2) {.smaller}

Some words in our dataset have a `Frequency` of `0`. One strategy is to *add one* to those values before taking the log.

```{r mutate2}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_conc = df_conc %>%  ### piping!
  mutate(log_freq = log10(Frequency + 1))

summary(df_conc$log_freq)
```



## `group_by` specific factors and `summarise` {.smaller}

- The `group_by` function is extremely useful for **grouping** your data by specific factors.
- You can `%>%` this into `summarise` to calculate summary statistics for each group. 


```{r groupby}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_conc %>%
  group_by(Dom_Pos) %>%
  summarise(mean_conc = mean(Concreteness)) %>%
  head(2)
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

In addition to calculating `mean`, try also calculating the `sd`; additionally, try calculating the `mean` and `sd` for `log_freq` by `Dom_Pos`. 
:::
:::

## Counting the number of observations {.smaller}

Often relevant for a `group_by` calculation is the *number* of observations per group.

```{r groupby_count}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_conc %>%
  group_by(Dom_Pos) %>%
  summarise(mean_conc = mean(Concreteness),
            count = n()) %>%
  head(2)
```

## Hands-on practice {.smaller}

Now let's get some hands-on practice applying these new techniques. 

::: {.incremental}
- Read in the `pokemon.csv` dataset from the [`ucsd_211_datasets` GitHub](https://github.com/seantrott/ucsd_css211_datasets) using `read_csv`. 
- Use `group_by` to *count* the number of Pokemon of each `Type 1`. Which `Type 1` is most frequent? 
- Then, `filter` the dataset to only include Pokemon of the most frequent type (based on what you learned). 
- Of this filtered dataset, `group_by` `legendary` status and calculate the `mean(Attack)`
:::


## Hands-on practice (solution) {.smaller}



```{r hands-on}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

### Part 1
df_pokemon <- read_csv("https://raw.githubusercontent.com/seantrott/ucsd_css211_datasets/main/main/wrangling/pokemon.csv")

### Part 2
max_type = df_pokemon %>% 
  group_by(`Type 1`) %>% 
  summarise(count = n()) %>%
  slice_max(count) %>%
  pull(`Type 1`)

### Parts 3-4
df_pokemon %>%
  filter(`Type 1` == max_type) %>%
  group_by(Legendary) %>%
  summarise(mean_attack = mean(Attack))
```


::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

Notice any functions we haven't discussed?
:::
:::

## Some new functions {.smaller}

Above, we used a combination of `read_csv`, `filter`, `group_by`, and `summarise`. But we also used some other new functions:

::: {.incremental}
- `slice_max`: returns rows with the **max** values of a given column. 
   - Can also use `slice_head`, `slice_tail`, `slice_min`, and `slice_sample`. 
- `pull`: returns the individual vector/value from a given column. 
:::


::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

Any questions before we move on to more `dplyr` functions and more aspects of the `tidyverse` more generally?
:::
:::

## `select` specific columns {.smaller}

> The `select` function can be used to **select** certain columns for further analysis.

This is especially useful if you want to `%>%` those columns into, say, a correlation matrix.


```{r select}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_conc %>%
  select(log_freq, Concreteness, Frequency) %>%
  cor()
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

Try building a correlation matrix of the numerical columns from the Pokemon dataset (e.g., `Attack`, `Defense`, etc.). Are there any conclusions you can draw?
:::
:::


## Part 2: More complex wrangling

Missing data, reshaping, and joining.


## Missing values {.smaller}

> Real-world data often has **missing values**, which are treated as `NA` ("Not Available") by R. 

There are a few relevant issues here:

::: {.incremental}
- First, you need to **identify** the presence of missing values. 
- Second, you need to **address** the issue. 
  - Ignore them? 
  - Remove any rows with missing values?
  - Try to impute their values?
:::

## Identifying missing values {.smaller}

We can use the `is.na` function to check for missing values.

```{r identify-missing}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

### Reading in dataset
df_titanic <- read_csv("https://raw.githubusercontent.com/seantrott/ucsd_css211_datasets/main/main/wrangling/titanic.csv")

### Check for missing values; returns a vector of boolean values
is.na(df_titanic$Age)[0:5]

```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

How many missing values are there total in the `Age` column? What about `Survived` or `Cabin`? (**Hint**: You can `sum` boolean values like `TRUE` and `FALSE`).
:::
:::

## Identify missing values (pt. 2) {.smaller}

We can combine some handy functions (`across` and `everything`) to sum up the number of `NA` values per column.

```{r identify-missing2}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

## Get number of NA across all columns
df_titanic %>% 
  summarise(across(everything(), ~ sum(is.na(.))))
```


::: {.fragment}
::: {.callout-note icon="false"}
`NA` values seem disproportionately located in the `Cabin` column. 
:::
:::

## Addressing missing values {.smaller}

Now that we've **identified** missing values, we have a few options:

::: {.incremental}
- We could *ignore* them, i.e., perform analyses as if they don't exist. 
- We could *remove* missing values from our dataset. 
- We could *impute* the missing values (guess). 
:::

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

What do you think are the trade-offs of each approach?
:::
:::


## The problem with ignoring `NA` {.smaller}

Unsurprisingly, ignoring `NA` values doesn't mean they go away. 

- In some cases, analyses or visualizations may "quietly" ignore them. 
  - This is bad because it means you're not *fully aware* of what goes into an analysis.
- In other cases, it'll interfere with your summary statistics.

```{r na-mean}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

## Mean defaults to NA b/c of missing values
mean(df_titanic$Age)

## Temporary fix is to remove them from this calculation...
mean(df_titanic$Age, na.rm = TRUE)
```

## Removing missing values {.smaller}

Another common approach to `NA` values is simply to **remove them**. 

We can do this selectively in a given analysis, like below:

```{r na-mean1}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

## Temporary fix is to remove them from this calculation...
mean(df_titanic$Age, na.rm = TRUE)
```

Or we can actually `filter` our dataset to remove them entirely (or use `drop_na`).

```{r na-mean2}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_titanic_subset = df_titanic %>%
  filter(!is.na(Age))
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

Do you think we should *always* just remove all `NA` values? Why or why not?
:::
:::

## Not all `NA` values matter equally {.smaller}

Whether or not we *care* that something is `NA` depends on our research questions!

- `Age` might be relevant for analyses of `Survived`. 
- `Cabin` may not really matter. 

```{r na-cabin}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

### Cabin mostly NA, but maybe that's fine?
sum(is.na(df_titanic$Cabin))

df_titanic$Cabin[1:10]
```

## How do `NA` values distribute? {.smaller}

One factor that might affect our approach is how `NA` values **distribute**.

- Are they distributed randomly with respect to some other variable (e.g., `Survived`)? 
- Or are they *systematic*? 
- Either way, it's helpful to know.


```{r na-survived}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

### Lower NA information about Cabins for people who survived
df_titanic %>%
  group_by(Survived) %>%
  summarise(proportion_na = mean(is.na(Cabin)))
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

What is the relative rate of `NA` values across levels of `Survived` for `Age`? 
:::
:::

## Imputing missing values {.smaller}

> **Imputing** values means *inferring* them based on some other set of properties or set of assumption. I.e., an "educated guess". 

- Can set to `mean` or `median` value (e.g., `mean(Age`)). 
- Can "guess" value based on some other property (e.g., `mean(Age)` for that `Pclass`).


```{r na-impute}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_titanic = df_titanic %>%
  group_by(Pclass) %>% ### Group by passenger class
  mutate(Age_imputed = ifelse(
    is.na(Age), mean(Age, na.rm = TRUE),  ### Replace with mean for that Pclass
    Age)) %>% ### otherwise just use original age
  ungroup()


df_titanic %>%
  group_by(Pclass) %>%
  summarise(mean_age = mean(Age_imputed))
```


## Missing values: wrap-up {.smaller}


::: {.incremental}
- Real-world data often has **missing values**.
- At minimum, it's important to *identify* and *characterize* `NA` values in your dataset.
- Ideally, you should develop a *strategy* for dealing with them with intentionality.
:::

::: {.fragment}
::: {.callout-note icon="false"}
Sometimes missing values show up when we **join** datasets. We'll discuss joining soon.
:::
:::


## Reshaping data {.smaller}

> **Reshaping data** means converting it to either a "longer" format (from a "wide" format) or the other way around. 

::: {.incremental} 
- In many cases, this means making data **tidy** (long format). 
- The `tidyr` package contains key functions for this, such as `pivot_longer` and `pivot_wider`. 
- `tidyr` also includes functions to `separate` information from a single column across multiple columns (or `unite`, to do the opposite.)
:::

::: {.fragment}
::: {.callout-note icon="false"}
The best way to understand this is to look at some examples!
:::
:::

## `pivot_longer`: from wide to long {.smaller}

Let's start by creating a **wide** dataset, which might be considered "messy".

```{r pivot_longer1}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_messy_1 = data.frame(john = c(10, 11),
                        mary = c(20, 25),
                        jane = c(5, 10),
                        treatment = c('a', 'b'))
df_messy_1
```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

Why is this considered "messy"? And what would a "long" version look like?
:::
:::

## `pivot_longer`: from wide to long {.smaller}

We can use the `pivot_longer` function to transform this into a **long** format.

```{r pivot_longer2}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_messy_1 %>%
  pivot_longer(cols = c(john, mary, jane), ### which columns to pivot
               names_to = 'name', ### what to call column with keys
               values_to = 'result') ### what to call column with values
```

::: {.incremental} 
- Crucial argument is `cols` (which columns to pivot).
- `names_to` and `values_to` are helpful for making the resulting table more readable.
:::


## `pivot_longer`: from wide to long {.smaller}

Let's look at another example.

```{r pivot_longer3}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_messy_2 = data.frame(name = c('john', 'mary', 'jane'),
                        a = c(10, 20, 5),
                        b = c(11, 25, 10))

df_messy_2

df_tidy = df_messy_2 %>%
  pivot_longer(cols = c(a, b),
               names_to = "treatment",
                values_to = "result")

df_tidy
```

## `pivot_wider`: back to wide format! {.smaller}

> `pivot_wider` reshapes data into **wide** format.

```{r pivot_wider}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

### To recreate first table
df_tidy %>%
  pivot_wider(names_from = "name",
              values_from = "result")

### To recreate second table
df_tidy %>%
  pivot_wider(names_from = "treatment",
              values_from = "result")
```


## Your turn! (pt. 1) {.smaller}

If you've loaded `tidyverse`, you should have access to (at least) the following data tables:

- `table1` 
- `table2` 

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

How would you `pivot_wider` `table2` such that it resembles `table1`?
:::
:::

## Your turn! (pt. 2) {.smaller}

Now let's try `pivot_longer` with a more complex example. Load the **missing work** example from [this Full Stack Economics graph](https://www.fullstackeconomics.com/p/18-charts-that-explain-the-american-economy). 


```{r pivot_practice}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

### Part 1
df_work <- read_csv("https://raw.githubusercontent.com/seantrott/ucsd_css211_datasets/main/main/wrangling/missing_work.csv")
head(df_work, 3)


```


::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

Right now, the table stores different *reasons for missing work* as different *columns*. 

Instead, let's `pivot_longer` so the table has a column for `Reason` (for missing) and a column for `Count` (number of days missed). 

:::
:::


## Other handy `tidyr` functions {.smaller}

Although **pivoting** is probably the *main* function enabled by `tidyr`, it also contains other valuable reshaping functions:

::: {.incremental} 
- `separate` "splits" values within a given column into multiple columns. 
- `unite` does the opposite.
:::



```{r separate_example}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

table3 %>%
  separate(rate,  ### column to separate
           into = c("cases", "population"),  ### columns to split into
           convert = TRUE, ### convert into numeric
           sep = "/") ### separator currently separating values

```

## Joining data {.smaller}

> **Joining** is an operation in which columns from data `Y` are added to data `X`, matching observations based on shared variables (or "keys"). 

::: {.incremental} 
- In real-world analysis, data is often stored across multiple sources. 
- *Joining* brings together related information from separate tables. 
- Key types: `left_join()`, `right_join()`, `inner_join()`, `full_join()`
:::

::: {.fragment}
**Example scenario:**

- Table 1: State election results (`state`, `biden_pct`, `turnout`)  
- Table 2: State demographics (`state`, `median_income`, `college_pct`)

Here, we might **join** the tables based on the overlapping `state` variable.

**Goal**: Analyze how `median_income` and `college_pct` relates to voting patterns.
:::


## Types of joins {.smaller}

Given tables `X` and `Y`, you have several options for how to **join** them.

::: {.incremental} 
- `inner_join`: only keep observations from `X` with matching key in `Y`. 
- `left_join`: keep all observations from `X`, join matching ones from `Y`.
- `right-join`: keep all observations from `Y`, join matching ones from `X`.
- `full_join`: keep all observations in both `X` and `Y`. 
:::

## Simple join example {.smaller}

Suppose we have two tables: `table4a` contains information about the **cases** in each country (with different columns for different years), and `table4b` contains information about the **population** of each country (again, with different columns for different years).

Ideally, we'd like to *join* these tables.

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

What steps will we need to take to *join* them?

(**Hint**: It may involve more than joining.)
:::
:::

## Simple join example {.smaller}

```{r join_example}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

### First, tidy each table
tidy4a = table4a %>%
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b = table4b %>%
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")

### Then, join
left_join(tidy4a, tidy4b)

```

::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

In this case, would it matter whether we use `left_join` or `inner_join`? If not, under what circumstances would this distinction matter?
:::
:::

## A more complex example {.smaller}

We've already loaded information about word **concreteness**. Now let's load another dataset about the average **age of acquisition** at which words are learned, and *combine* that with the concreteness dataset.

```{r aoa}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

df_aoa <- read_csv("https://raw.githubusercontent.com/seantrott/ucsd_css211_datasets/main/main/wrangling/AoA.csv")
nrow(df_aoa)
head(df_aoa)

```


::: {.fragment}
::: {.callout-note icon="false"}
### 💭 Check-in

How would you **join** this dataset with the concreteness dataset? What kind of *join* operation might you want to use?
:::
:::

## A more complex example (pt. 2) {.smaller}

The concreteness and AoA datasets are not **fully-overlapping**. Concreteness contains words not in AoA and vice versa.


```{r aoa_join}
#| echo: true
#| code-line-numbers: true
#| output-location: fragment

### Inner join (keep only intersection)
df_inner = df_conc %>%
  inner_join(df_aoa)
nrow(df_inner)

### Left join (keep all concreteness)
df_left = df_conc %>%
  left_join(df_aoa)
nrow(df_left)

### Right join (keep all AoA)
df_right = df_conc %>%
  right_join(df_aoa)
nrow(df_right)

### full join (keep everything)
df_full = df_conc %>%
  full_join(df_aoa)
nrow(df_full)

```

## Joining: the details {.smaller}

When **joining**, there are a few other details we've glossed over:

- `by`: this specifies which *keys* to join on (will default to all overlapping). 
- `multiple`: how to handle cases where rows from `X` match multiple rows from `Y`. 
- `unmatched`: how to handle cases where rows would be dropped. 



## Putting it all together {.smaller}

These lectures have covered the basics of **data wrangling** with the `tidyverse`. 

A typical workflow will involve:

::: {.incremental}
- **Reading** data (`read_csv`). 
- **Exploring** the data (`is.na`, `summary`, `summarise`). 
- **Clean** data (`mutate`, `filter`). 
- **Reshape** if needed `(pivot_longer/wider`). 
- **Join** if needed (`inner_join`, etc.). 
:::

::: {.fragment}
::: {.callout-note icon="false"}
The end result is a dataset ready for further *analysis* and *visualization* (coming up!). 
:::
:::


